---
title: "Smoking Status and Self-reported Health in New York State"
author: "Puying Li"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
---

## Introduction

This report explores the association between smoking status and self-reported general health among adults in New York State.

## Research Question and Background

**Research question:**  
Is smoking status associated with self-reported general health among adults in New York State?

**Scientific Plausibility and Context: **

Smoking is a major risk factor for many chronic diseases, including cardiovascular disease, cancer, and respiratory illness (U.S. Department of Health and Human Services, 2014). These conditions can negatively affect physical functioning and quality of life, which may lead individuals who smoke to perceive their overall health as worse than non-smokers. Research has also shown that long-term smoking is associated with increased mortality and adverse health outcomes (Inoue-Choi et al., 2017). Additionally, self-reported health is widely used as an indicator of general well-being and has been shown to predict morbidity and mortality (Idler & Benyamini, 1997). Therefore, it is scientifically plausible that smoking status is associated with differences in self-reported health among adults.

**Refences:**

Idler, E. L., & Benyamini, Y. (1997). Self-rated health and mortality: a review of twenty-seven community studies. Journal of health and social behavior, 38(1), 21–37.

Inoue-Choi, M., Liao, L. M., Reyes-Guzman, C., Hartge, P., Caporaso, N., & Freedman, N. D. (2017). Association of Long-term, Low-Intensity Smoking With All-Cause and Cause-Specific Mortality in the National Institutes of Health-AARP Diet and Health Study. JAMA internal medicine, 177(1), 87–95. https://doi.org/10.1001/jamainternmed.2016.7511

National Center for Chronic Disease Prevention and Health Promotion (US) Office on Smoking and Health. The Health Consequences of Smoking—50 Years of Progress: A Report of the Surgeon General. Atlanta (GA): Centers for Disease Control and Prevention (US); 2014. Available from: https://www.ncbi.nlm.nih.gov/books/NBK179276/

## Data Preparation

Data from the New York State BRFSS 2023 survey were used for this analysis. Two variables were selected: smoking status (_RFSMOK3) and self-reported general health (GENHLTH). The smoking variable was recoded into a new variable called “SMOKE,” and only respondents who answered “Yes” or “No” were included in the analysis. Responses such as “Don’t know,” “Refused,” or missing values were excluded. 
The general health variable was categorized into five levels: Excellent, Very good, Good, Fair, and Poor. Both variables were converted into categorical (factor) variables to facilitate summary statistics and visualization. The final dataset included only participants with complete data on both variables.

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(readr)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r load-data}
dat <- read_csv("NYSDOH_BRFSS_SurveyData_2023_1.csv", show_col_types = FALSE)
```

```{r data-prep}
analysis_data <- dat %>%
  dplyr::select(`_RFSMOK3`, GENHLTH) %>%
  dplyr::rename(SMOKE = `_RFSMOK3`) %>%
  dplyr::filter(!is.na(SMOKE), !is.na(GENHLTH), SMOKE %in% c("Yes", "No")) %>%
  dplyr::mutate(
    Smoking = factor(SMOKE, levels = c("No", "Yes")),
    Health  = factor(GENHLTH, levels = c("Excellent", "Very good", "Good", "Fair", "Poor"))
  )
```


## Grouped Summary Statistics

A cross-tabulation of smoking status and self-reported general health was created to examine differences between smokers and non-smokers. Among non-smokers, a larger proportion reported better health, with many respondents indicating “Excellent” or “Very good” health. In contrast, smokers were less likely to report excellent health and more likely to report fair or poor health.
For example, only a small percentage of smokers reported “Excellent” health compared to non-smokers, while the proportion reporting “Fair” or “Poor” health was higher among smokers. These patterns suggest that smoking status may be associated with worse self-reported health outcomes.
Overall, the descriptive statistics indicate a potential negative relationship between smoking and perceived general health in adults in New York State.


```{r}

tab_counts <- table(analysis_data$Smoking, analysis_data$Health)
tab_counts

```
```{r}
round(prop.table(tab_counts, margin = 1), 3)

```
```{r}
ggplot(analysis_data, aes(x = Smoking, fill = Health)) +
  geom_bar(position = "fill") +
  ylab("Proportion") +
  ggtitle("Self-reported Health by Smoking Status (NYS BRFSS 2023)") +
theme_minimal(base_size = 12)

```

```{r}
ggsave(
  "health_smoking_plot.pdf",
  width = 7,
  height = 5
)

```


## Comparative Visualization
The stacked bar chart illustrates differences in self-reported health by smoking status among adults in New York State. Non-smokers show higher proportions of “Excellent” and “Very good” health, while smokers have relatively larger proportions reporting “Fair” or “Poor” health.
These patterns suggest that smoking is associated with worse perceived health outcomes. The visualization supports the descriptive statistics and indicates a potential negative relationship between smoking and general health status.
Overall, the results are consistent with existing evidence that tobacco use adversely affects physical health and well-being.

## AI Use Disclosure Statement

Yes, I used ChatGPT to assist with understanding concepts and debugging R code that I wrote myself. The final analysis, interpretation, and written content were completed independently by me.

