---
title: "VTPEH 6270 - Check Point 02"
author: "Puying Li"
date: "2026-02-05"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)

brfss <- readr::read_csv("NYSDOH_BRFSS_SurveyData_2023.csv", show_col_types = FALSE)
dim(brfss)


```


```{r subset_vars}
vars_keep <- c("AGE", "SEXVAR", "BPHIGH6", "BPMEDS1", "_BMI5CAT", "_BMI5", "_LLCPWT")


df <- brfss %>%
  dplyr::select(dplyr::all_of(vars_keep))

glimpse(df)

```
```{r clean_vars}
df_clean <- df %>%
  dplyr::rename(
    age_group = AGE,
    sex = SEXVAR,
    bp_high = BPHIGH6,
    bp_meds = BPMEDS1,
    bmi_cat = `_BMI5CAT`,
    weight = `_LLCPWT`
  ) %>%
  dplyr::filter(
    !stringr::str_detect(age_group, "Refused|Missing|Not sure|Don't know"),
    !stringr::str_detect(sex, "Refused|Missing|Not sure|Don't know"),
    !stringr::str_detect(bp_high, "Refused|Missing|Not sure|Don't know|Not asked"),
    !stringr::str_detect(bp_meds, "Refused|Missing|Not sure|Don't know|Not asked"),
    !stringr::str_detect(bmi_cat, "Refused|Missing|Not sure|Don't know")
  )

glimpse(df_clean)

```


```{r variable_overview}
variable_overview <- tibble(
  Variable = names(df_clean),
  Type = if_else(sapply(df_clean, is.numeric), "Continuous", "Categorical"),
  Class = sapply(df_clean, function(x) paste(class(x), collapse = ", "))
)

variable_overview
```



```{r bmi_frequency}
df_clean %>%
  count(bmi_cat, sort = TRUE)

```

```{r bp_frequency}
df_clean %>%
  count(bp_high, sort = TRUE)
```


```{r meds_frequency}
df_clean %>%
  count(bp_meds, sort = TRUE)

```
Most individuals with high blood pressure reported taking blood pressure medication, although a substantial portion were not currently using medication.


```{r sex_frequency}
df_clean %>%
  count(sex, sort = TRUE)
```

```{r}
df_clean %>% count(age_group, sort = TRUE)

```

```{r bmi5_frequency}
df_clean %>%
  count(`_BMI5`, sort = TRUE)
```

```{r weight_summary}
summary(df_clean$weight)
```
Body weight showed a right-skewed distribution, with a mean weight higher than the median, indicating the presence of some extremely high values in the dataset.

```{r table_bmi}
df_clean %>% count(bmi_cat, sort = TRUE)

```

```{r bmi_plot}
ggplot(df_clean, aes(x = bmi_cat)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "BMI Category Distribution among Adults with High Blood Pressure",
    x = "BMI Category",
    y = "Count"
  )
```
The majority of adults with high blood pressure were classified as either obese or overweight. Obese individuals represented the largest group, followed closely by those who were overweight. Normal weight individuals were less common, while underweight individuals made up a very small proportion of the sample. This suggests a strong association between elevated body weight and high blood pressure in this population.

```{r weight_histogram}
library(ggplot2)

mean_w <- mean(df_clean$weight, na.rm = TRUE)
median_w <- median(df_clean$weight, na.rm = TRUE)

ggplot(df_clean, aes(x = weight)) +
  geom_histogram(aes(fill = "Weight"), bins = 30, color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = mean_w, color = "Mean"), linewidth = 1) +
  geom_vline(aes(xintercept = median_w, color = "Median"), linewidth = 1, linetype = "dashed") +
  scale_color_manual(values = c("Mean" = "blue", "Median" = "red")) +
  labs(
    x = "Weight",
    y = "Frequency",
    title = "Histogram of Weight",
    caption = "Distribution of weight among survey participants"
  ) +
  theme_minimal()
```
The histogram shows the distribution of weight among survey participants with high blood pressure. Most observations are concentrated at lower weight values, with a long right tail indicating a right-skewed distribution. The mean weight is higher than the median, suggesting the presence of extreme high-weight outliers. This implies that while most individuals fall within a moderate weight range, a small number of participants have very high weights.

```{r scatter_plot}
ggplot(df_clean, aes(x = age_group, y = weight)) +
  geom_jitter(alpha = 0.5, width = 0.2, color = "steelblue") +
  labs(
    title = "Scatter Plot of Weight by Age Group",
    x = "Age Group",
    y = "Weight"
  ) +
  theme_minimal()

```
The scatter plot illustrates the relationship between weight and age group among individuals with high blood pressure. Each point represents an individual’s weight within a specific age category. The spread of weight values increases in older age groups, particularly among those aged 55 and above. This suggests greater variability in body weight among older adults, with several high-weight outliers appearing in the older age categories.

•	Did you use AI? Yes
•	If yes: Write a short disclosure statement (2–3 sentences) describing:
•	Which tool(s) you used.ChatGPT
•	How you used it (e.g., to help decide on an analysis approach, to generate a first draft of code, to improve or debug code you had written yourself, etc.).
Example: “This document was generated using Claude to generate original code, which was then reviewed and adjusted” or “This document was generate using ChatGPT to assist with debugging of code generated by the author”.
Please keep your statement brief and honest. The goal is transparency, not detail: we do not need exact prompts or transcripts, just a clear sense of how AI supported this work.
This document was generated using ChatGPT to assist with data cleaning, exploratory analysis, and debugging of R code written by myself. The AI tool was used to help structure visualizations and resolve coding errors, while all final decisions and interpretations were made by myself.



